Non-convergence in iterative imputation

### 2025-04-16 Manuscript/manuscript.qmd
update sim design

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(mice)
library(ggmice)
library(ggplot2)
```

# Abstract (old)

@@

@@

We investigate non-convergence of the MICE algorithm through model-based simulation in `R` [version 3.6.3; @R]. The number of simulation repetitions is 1000, and the simulation set-up is summarized in the pseudo-code below. The complete `R` script of the simulation study is available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).

```
# pseudo-code of simulation 
1. simulate data 
for (number of simulation runs from 1 to 1000)
 for (proportion of missing cases between 5% and 95%)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. perform analysis of scientific interest
   5. compute non-convergence diagnostics 
   6. pool results across imputations
   7. compute performance measures
 8. combine outcomes of all conditions
9. aggregate outcomes across simulation runs

```
### Data-generating mechanism.

In this study, sampling variance is not of interest. Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions [@vink14]. The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem. Let the predictor space be defined as 
$$
\begin{aligned}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 \\
4 & 16 & 4.8\\
1.8 & 4.8 & 9
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{aligned}
$$
A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm]. Subsequently, a fourth variable is constructed as outcome variable $Y$. For each unit $i = 1, 2,..., N$, let 
$$
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
$$
where $\epsilon \sim \mathcal{N}(0, 100)$. This results in a dataset $y$, where $y_{\mathrm mis} = \emptyset$. In other words, $y$ contains the complete set of observations for all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$). From the complete dataset we obtain the true values of the scientific estimands $Q$.

@@

We consider four types of $Q$s that are often of interest in empirical research. Namely, two descriptive statistics: the mean $\mu_j$ and standard deviation $\sigma_j$ of each variable $j$. We also consider the regression coefficients of the predictors: $\beta_1$, $\beta_2$, and $\beta_3$ in regression equation
$$Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,$$
where $Y'$ is the expected value of the outcome. The final estimand is the proportion of variance explained by the set of predictors: coefficient of determination $r^2$ (note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$). The scientific estimands are estimated by $\bar{Q}$ in each simulation condition.

@@

In each simulation repetition, we *ampute* the complete dataset conform the missingness conditions. The proportion of incomplete cases $p_{\mathrm inc}$ varies between 5% and 95%, in incremental steps ($p_{\mathrm inc}$ = .05, .25, .50, .75, .95). We consider all possible univariate and multivariate patterns of missingness, conform a 'missing completely at random' missingness mechanism [MCAR; @rubin87]. I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on $p_{\mathrm inc}$. We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm obs}, y_{\mathrm mis}$} for each of the five missingness conditions.

@@

Subsequently, we obtain the completed data per imputation, {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice]. We estimate univariate $Q$s directly from the completed data, by averaging across imputations to obtain $\bar{Q}$. Estimates of multivariate $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R]. We pool the resulting estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform @vink14.

>>>>

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(mice)
library(ggmice)
library(ggplot2)
library(dplyr)
load("../results/out.RData")
source("../R/evaluate_estimates.R")
performance <- evaluate_est(out)
performance[performance$term == "Y", "term"] <- "b0"
results <- # simulation results split by condition and regression coefficient
performance |> 
  group_by(method, mech, prop, .it, term) |> 
  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), \(x) mean(x, na.rm = TRUE))) |>
  mutate(mech = factor(mech, levels = c("MCAR", "MAR"), ordered = TRUE),
         prop = factor(prop),
         term = factor(term))
```

```{r, echo=FALSE}
# # plot results per sim rep
# performance |>
#   filter(term == "X1", mech == "MAR", method == "MICE", prop == 0.75) |>
#   ggplot(aes(x = .it, y = bias)) + #, color = as.factor(prop)
#   geom_line(aes(group = .sim)) +
#   geom_smooth(method = "loess") +
#   # labs(
#   #   title = "Bias regression coefficient",
#   #   x = "Iteration number",
#   #   y = "Bias"
#   # ) +
#   theme_classic()
```

@@

```{r}
results |>
  filter(method == "MICE") |> #, prop == 0.75 mech == "MAR",
  ggplot(aes(x = .it, y = bias, color = prop)) +
  geom_hline(yintercept = 0, alpha = 0.1) + 
  geom_line() +
  facet_grid(term ~ mech) +
  theme_classic()

results |>
  filter(method == "MICE") |> #, prop == 0.75 mech == "MAR",
  ggplot(aes(x = .it, y = psrf_mean, color = prop)) +
  geom_hline(yintercept = 1, alpha = 0.1) + 
  geom_line() +
  facet_grid(term ~ mech) +
  theme_classic()

@@

We investigate non-convergence of the MICE algorithm through model-based simulation in `R` [version 4.4.2; @R]. The number of simulation repetitions is 2000, and the simulation set-up is summarized in the pseudo-code below. The complete `R` script of the simulation study is available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint) [TODO: make Zenodo DOI].

```
# pseudo-code of simulation 
for (number of simulation runs) {
  1. simulate complete data 
  for (missingness conditions) {
    2. create missingness
    for (number of iterations) {
      3. impute missingness
      4. estimate quantities of scientific interest
      5. apply performance measures to the estimates
      6. compute non-convergence diagnostics
    }
  }
7. aggregate outcomes across simulation runs
```

### Data-generating mechanism.

The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem. Let the predictor space be defined as

$$
\begin{aligned}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
0\\
0\\
0
\end{pmatrix}\!\!,
\begin{pmatrix}
1 & 0.5 & 0.5 \\
0.5 & 1 & 0.5 \\
0.5 & 0.5 & 1
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{aligned}
$$

In each simulation repetition, a finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm]. Subsequently, a fourth variable is constructed as outcome variable $Y$. For each unit $i = 1, 2,..., N$, let

$$
Y_i = X_{1i} + X_{2i} + X_{3i} + \epsilon_i ,
$$
where $\epsilon \sim \mathcal{N}(0, 1)$. This results in a dataset $y$, where $y_{\mathrm mis} = \emptyset$. In other words, $y$ contains the complete set of observations for all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$).

@@

We consider four multivariate quantities of scientific interest ($Qs$). Namely, the regression coefficients of the analysis model,

$$Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,$$
where $Y'$ is the expected value of the outcome. From the data-generating mechanism, we obtain the true values of the scientific estimands: $\beta_0 = 0$, $\beta_1 = \beta_2 = \beta_3 = 1$. The scientific estimands are estimated by $\bar{Q}$ in each simulation condition.

<!-- The final estimand is the proportion of variance explained by the set of predictors: coefficient of determination $r^2$ (note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$). -->

@@

We consider two sets of simulation conditions, 'missingness conditions' and 'early stopping' conditions. The simulation study is a fully factorial design, as visualized in Table XYZ.

In each simulation repetition, we *ampute* the complete dataset conform six missingness conditions: two missing data mechanisms and three proportions of incomplete cases. The missing data mechanisms are 'missing completely at random' [MCAR; @rubin87], where the probability to be missing is the same for all $N \times k$ cells in $y$, and a right-tailed 'missing at random' (MAR) mechanism, where the probability to be missing is a function of the observed data and higher values are more likely to be missing. The proportion of incomplete cases $p_{\mathrm inc}$ is set to 25%, 50%, and 75%.

<!-- Table -->
<!-- mechanisms: MCAR, MAR -->
<!-- proportion of incomplete cases: 25%, 50%, 75% -->

<!-- \begin{table}[] -->
<!-- \begin{tabular}{lllll} -->
<!-- \hline -->
<!-- Missingness &  &                 &  & Early stopping            \\ \cline{1-3} \cline{5-5}  -->
<!-- Mechanism   &  & Proportion      &  & Iteration number          \\ \cline{1-1} \cline{3-3} \cline{5-5}  -->
<!-- MCAR, MAR   &  & 0.25, 0.5, 0.75 &  & 1, 2, 3, ..., 98, 99, 100 \\ \hline -->
<!-- \end{tabular} -->
<!-- \end{table} -->

$$
\begin{aligned}
\hline
&\mathrm{Missingness \ conditions} & & & &\mathrm{Early \ stopping \ conditions} \\
\hline
&\begin{matrix}
\mathrm{MCAR} \\
\mathrm{MAR \ } 
\end{matrix}
\quad
&\times
\quad
&\begin{matrix}
25\% \\
50\% \\
75\%
\end{matrix}
\quad
&\times
\quad
&\begin{matrix}
1 \\
2 \\
... \\
99 \\
100 \\
\end{matrix} \\
&\mathrm{mechansim} &&\mathrm{incomplete \ cases} &&\mathrm{iteration(s)} \\
\hline
\end{aligned}
$$

We consider all possible multivariate patterns of missingness, as visualized in Figure \ref{fig:patterns}.

```{r patterns, echo=FALSE, eval=FALSE}
source("../R/generate_datasets.R")
pat <- create_patterns(3)
d <- dat[1:10,]
is.na(d) <- pat == 1
d
ggmice::plot_pattern(d)
```

We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\mathrm obs}, y_{\mathrm mis}$} for each of the six missingness conditions.

@@

Subsequently, we obtain the completed data per imputation, {$y_{\mathrm obs}, \dot y_{\mathrm imp, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice]. Estimates of scientific quantities $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R]. We pool the resulting $m$ estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform Rubin's rules [function `mice::pool()`; @mice].
